# Graphs

By default, it is possible to make a lot of graphs with R without the need of any external
packages. However, in this chapter, we are going to learn how to make graphs using `ggplot2` which
is a very powerful package that produces amazing graphs. There is an entry cost to `ggplot2` as it
works in a very different way than what you would expect, especially if you know how to make plots
with the basic R functions already. But the resulting graphs are well worth the effort and once
you'll know more about `ggplot2` you'll see that it in a lot of situations it is actually faster
and easier.

## Resources

Before showing some examples and the general functionality of `ggplot2`, I list here some online
resources that I keep coming back to:

* [R Graphics Cookbook](http://www.cookbook-r.com/Graphs/)

* [R graph gallery](http://www.r-graph-gallery.com/portfolio/ggplot2-package/)

* [Tufte in R](http://motioninsocial.com/tufte/)

* [ggplot2 extensions](http://www.ggplot2-exts.org/gallery/)

* [ggthemes Vignette](https://cran.r-project.org/web/packages/ggthemes/vignettes/ggthemes.html)

* [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)

I have a cookbook approach to using `ggplot2`; I try to find an example online that looks similar
to what I have in mind, copy and paste the code and then adapt it to my case. The above resources
are the ones I consult the most in these situations (I also go back to past code I've written, of
course). Don't hesitate to skim these resources for inspiration and to learn more about some
extensions to `ggplot2`. In the next subsections I am going to show you how to draw the most common
plots, as well as show you how to customize your plots with `ggthemes`.


## Examples

### Barplots

To follow the examples below, load the following libraries:

```{r}
library(ggplot2)
library(ggthemes)
```

`ggplot2` is an implementation of the *Grammar of Graphics* by @wilkinson2006, but you don't need
to read the books to start using it. If we go back to the Star Wars data (contained in `dplyr`),
and wish to draw a barplot of the gender, the following lines are enough:


```{r}
data(starwars)

ggplot(starwars, aes(gender)) +
  geom_bar()
```

The first argument of the function is the data (called `starwars` in this example), and then the
function `aes()`. This function is where you list the variables you want to map, and to quote the
help file of `aes()`, *describes how the variables are mapped to visual properties (aesthetics) of
geoms*. You can get different kind of plots by using different `geom_` functions.


### Density

`geom_density()` is the *geom* that allows you to get density plots:

```{r}
ggplot(starwars, aes(height)) +
  geom_density()
```

Let's go into more detail now; what if you would like to plot the densities for females and males
only (removing the droids from the data first)? This can be done by first filtering the data using
`dplyr` and then separating the dataset by gender:


```{r, eval=FALSE}
starwars %>%
  filter(gender %in% c("female", "male"))
```

The above lines do the filtering; only keep gender if gender is in the vector `"female", "male"`.
This is much easier than having to write `gender == "female" | gender == "male"`. Then, we pipe
this dataset to `ggplot`:

```{r}
starwars %>%
  filter(gender %in% c("female", "male")) %>%
  ggplot(aes(height, fill = gender)) +
  geom_density()
```

Let's take a closer look to the `aes()` function: I've added `fill = gender`. This means that the
there will be one density plot for each gender in the data, and each will be colored accordingly.
This is where `ggplot2` might be confusing; there is no need to write explicitly (even if it is
possible) that you want the *female* density to be red and the *male* density to be blue. You just
map the variable `gender` to this particular aesthetic. You conclude the plot by adding
`geom_density()` which is this case is the plot you want. We will see late on how to change the
colors of your plot.


### Line plots

For the line plots, we are going to use official unemployment data (the same as in the previous
chapter, but with all the available years). Get it from
[here](https://github.com/b-rodrigues/r_for_statec/tree/master/datasets/unemployment/all)
(downloaded from:
http://www.statistiques.public.lu/stat/TableViewer/tableView.aspx?ReportId=12950&IF_Language=eng&MainTheme=2&FldrName=3&RFPath=91).

Let's plot the unemployment for the canton of Luxembourg only:

```{r}
unemp_lux_data = import("datasets/unemployment/all/unemployment_lux_all.csv")

unemp_lux_data %>%
  filter(division == "Luxembourg") %>%
  ggplot(aes(year, unemployment_rate_in_percent, group = 1)) +
  geom_line()
```

Because line plots are 2D, you need to specify the y and x axes. There is also another option you
need to add, `group = 1`. This is to tell `aes()` that the dots have to be connected with a single
line. What if you want to plot more than one commune?

```{r}
unemp_lux_data %>%
  filter(division == "Luxembourg" | division == "Esch-sur-Alzette") %>%
  ggplot(aes(year, unemployment_rate_in_percent, group = division, color = division)) +
  geom_line()
```

This time, I've specified `group = division` which means that there has to be one line per as many
communes as in the variable `division`. I do the same for colors. I think the next example
illustrates how `ggplot2` is actually brilliant; if you need to add a third commune, there is no
need to specify anything else; no need to add anything to the legend, no need to specify a third
color etc:

```{r}
unemp_lux_data %>%
  filter(division %in% c("Luxembourg", "Esch-sur-Alzette", "Wiltz")) %>%
  ggplot(aes(year, unemployment_rate_in_percent, group = division, color = division)) +
  geom_line()
```

### Facets

```{r}
starwars %>%
  mutate(human = case_when(species == "Human" ~ "Human",
                           species != "Human" ~ "Not Human")) %>%
  filter(gender %in% c("female", "male"), !is.na(human)) %>%
  ggplot(aes(height, fill = gender)) +
  facet_grid(~human) +
  geom_density()
```


### Customization

Let's go back to the unemployment line plots.

```{r}
unemp_lux_data %>%
  filter(division %in% c("Luxembourg", "Esch-sur-Alzette", "Wiltz")) %>%
  ggplot(aes(year, unemployment_rate_in_percent, group = division, color = division)) +
  theme(legend.position = "bottom") +
  labs(title="Unemployment in Luxembourg, Esch/Alzette and Wiltz", x="Year", y = "Rate")+
  geom_line()
```

```{r}
unemp_lux_data %>%
  filter(division %in% c("Luxembourg", "Esch-sur-Alzette", "Wiltz")) %>%
  ggplot(aes(year, unemployment_rate_in_percent, group = division, color = division)) +
    theme_stata() +
    theme(legend.position = "bottom") +
  labs(title="Unemployment in Luxembourg, Esch/Alzette and Wiltz", x="Year", y = "Rate") +
  geom_line()
```

```{r}
unemp_lux_data %>%
  filter(division %in% c("Luxembourg", "Esch-sur-Alzette", "Wiltz")) %>%
  ggplot(aes(year, unemployment_rate_in_percent, group = division, color = division)) +
  theme_minimal() +
  scale_color_hc() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  labs(title="Unemployment in Luxembourg, Esch/Alzette and Wiltz", x = "Year", y = "Rate") +
  geom_line()
```
