# Graphs

By default, it is possible to make a lot of graphs with R without the need of any external
packages. However, in this chapter, we are going to learn how to make graphs using `ggplot2` which
is a very powerful package that produces amazing graphs. There is an entry cost to `ggplot2` as it
works in a very different way than what you would expect. But the resulting graphs are well worth
the effort.

## ggplot2

### Barplots

```{r}
library(ggplot2)
library(ggthemes)
```

```{r}
ggplot(starwars, aes(gender)) +
  geom_bar()
```

### Density

```{r}
ggplot(starwars, aes(height)) +
  geom_density()
```

```{r}
starwars %>%
  filter(gender %in% c("female", "male")) %>%
  ggplot(aes(height, fill = gender)) +
  geom_density()
```

```{r}
starwars %>%
  filter(gender %in% c("female", "male")) %>%
  ggplot(aes(height, fill = gender)) +
  geom_density(alpha = 0.5)
```

### Line plots

```{r}
unemp_lux_data = import("datasets/unemployment/all/unemployment_lux_all.csv")

unemp_lux_data %>%
  filter(division == "Luxembourg") %>%
  ggplot(aes(year, unemployment_rate_in_percent)) +
  geom_line()
```
