# Statistical models

In this chapter, we will not learn about all the models out there that you may or may not need.
Instead, I will show you how can use what you have learned until now and how you can apply these
concepts to modelling. Also, as you read in the beginning of the book, R has many many packages. So
the model you need is most probably already implemented in some package.

## Simple modelling

Suppose you have a variable `y` that you wish to explain using a set of other variables `x1`, `x2`,
`x3`, etc. Let's take a look at the `Housing` dataset from the `Ecdat` package:

```{r, include=FALSE}
library(Ecdat)

data(Housing)
```

```{r, eval=FALSE}
library(Ecdat)

data(Housing)
```

You can read a description of the dataset by running:

```{r, eval=FALSE}
?Housing
```


```
Housing                 package:Ecdat                  R Documentation

Sales Prices of Houses in the City of Windsor

Description:

     a cross-section from 1987

     _number of observations_ : 546

     _observation_ : goods

     _country_ : Canada

Usage:

     data(Housing)

Format:

     A dataframe containing :

     price: sale price of a house

     lotsize: the lot size of a property in square feet

     bedrooms: number of bedrooms

     bathrms: number of full bathrooms

     stories: number of stories excluding basement

     driveway: does the house has a driveway ?

     recroom: does the house has a recreational room ?

     fullbase: does the house has a full finished basement ?

     gashw: does the house uses gas for hot water heating ?

     airco: does the house has central air conditioning ?

     garagepl: number of garage places

     prefarea: is the house located in the preferred neighbourhood of the city ?

Source:

     Anglin, P.M.  and R.  Gencay (1996) “Semiparametric estimation of
     a hedonic price function”, _Journal of Applied Econometrics_,
     *11(6)*, 633-648.

References:

     Verbeek, Marno (2004) _A Guide to Modern Econometrics_, John Wiley
     and Sons, chapter 3.

     Journal of Applied Econometrics data archive : <URL:
     http://qed.econ.queensu.ca/jae/>.

See Also:

     ‘Index.Source’, ‘Index.Economics’, ‘Index.Econometrics’,
     ‘Index.Observations’
```

or by looking for `Housing` in the help pane of RStudio. Usually, you would take a look a the data
before doing any modelling:

```{r}
glimpse(Housing)
```

Housing prices depend on a set of variables such as the number of bedrooms, the area it is located
and so on. If you believe that housing prices depend linearly on a set of explanatory variables,
you will want to estimate a linear model. To estimate a *linear model*, you will need to use the
built-in `lm()` function:

```{r}
model1 = lm(price ~ lotsize + bedrooms, data = Housing)
```

`lm()` takes a formula as an argument, which defines the model you want to estimate. In this case,
I ran the following regression:

\[
\text{price} = \alpha + \beta_1 * \text{lotsize} + \beta_2 * \text{bedrooms} + \varepsilon
\]

where \(alpha, beta_1\) and \(beta_2\) are three parameters to estimate. To take a look at the
results, you can use the `summary()` method (not to be confused with `dplyr::summarise()`:

```{r}
summary(model1)
```
